### 2015-12-21 solo-ftrl.tex
Reword and reorganize some sentences
Polished grammar and typography.

The closest algorithm to a scale-free one in the OLO literature is the
\textsc{AdaGrad} algorithm~\cite{Duchi-Hazan-Singer-2011}.  It uses a
regularizer on each coordinate of the form
\begin{equation*}
R_t(w) = R(w) \left(\delta + \sqrt{\sum_{i=1}^{t-1} \|\ell_i\|_*^2} \right).
\end{equation*}
This kind of regularizer would yield a scale-free algorithm \emph{only} for
$\delta=0$.  Unfortunately, the regret bound in~\cite{Duchi-Hazan-Singer-2011}
becomes vacuous for such setting in the unbounded case. In fact, it requires
$\delta$ to be greater than $\|\ell_t\|_*$ for all time steps $t$, requiring
knowledge of the future (see Theorem~5 in~\cite{Duchi-Hazan-Singer-2011}). In
other words, despite of its name, \textsc{AdaGrad} is not fully adaptive to the
norm of the gradient vectors. Identical considerations hold for the
\textsc{FTRL-Proximal} in \cite{McMahan-Streeter-2010,McMahan-2014}: the
scale-free setting of the learning rate is valid only in the bounded case.

One simple approach would be to use a doubling trick on $\delta$ in order to
estimate on the fly the maximum norm of the losses. Note that a naive strategy
would still fail because the initial value of $\delta$ should be data-dependent
in order to have a scale-free algorithm. Moreover, we would have to upper bound
the regret in all the rounds where the norm of the current loss is bigger than
the estimate. Finally, the algorithm would depend on an additional parameter,
the ``doubling'' power. Hence, even in the case one would prove a regret bound,
%\footnote{For
%lack of space, we cannot include the regret bound for the doubling trick
%version. It would be exactly the same as in
%Theorem~\ref{theorem:regret-solo-ftrl}, following a similar analysis, but with
%the additional parameter of the doubling power.},
such strategy would give the
feeling that \textsc{FTRL} needs to be ``fixed'' in order to obtain a
scale-free algorithm.

@@

\begin{theorem}[Regret of \textsc{SOLO FTRL}]
\label{theorem:regret-solo-ftrl}
Suppose $K \subseteq V$ is a non-empty closed convex subset.  Let $D =
\sup_{u,v \in K} \|u - v\|$ be its diameter with respect to a norm $\|\cdot\|$.
Suppose that the regularizer $R:K \to \R$ is a non-negative lower
semi-continuous function that is $\lambda$-strongly convex with respect to
$\|\cdot\|$. The regret of \textsc{SOLO FTRL} satisfies
\begin{align*}
\Regret_T(u)
& \le \left( R(u) + \frac{2.75}{\lambda}\right) \sqrt{\sum_{t=1}^{T} \norm{\ell_t}_*^2}
+ 3.5 \min\left\{\frac{\sqrt{T-1}}{\lambda} , D\right\} \max_{t \le T} \|\ell_t\|_* \; .
\end{align*}
\end{theorem}

@@

\begin{corollary}[Regret Bound for Bounded Decision Sets]
\label{corollary:regret-solo-ftrl-bounded-set}
Suppose $K \subseteq V$ is a non-empty bounded closed convex subset.  Suppose
that $f:K \to \R$ is a non-negative lower semi-continuous function that is
$1$-strongly convex with respect to $\|\cdot\|$. \textsc{SOLO FTRL} with
regularizer
$$
R(w) = \frac{f(w)\sqrt{2.75}}{\sqrt{\sup_{v \in K} f(v)}}
\quad \text{satisfies} \quad
\Regret_T \le 13.3 \sqrt{\sup_{v \in K} f(v) \sum_{t=1}^{T} \norm{\ell_t}_*^2} \; .
$$
\end{corollary}

@@

The proof of Theorem~\ref{theorem:regret-solo-ftrl} relies on an inequality
(Lemma~\ref{lemma:useful}).  Related and weaker inequalities were proved by
~\cite{Auer-Cesa-Bianchi-Gentile-2002} and ~\cite{Jaksch-Ortner-Auer-2010}.
The main property of this inequality is that on the right-hand side $C$ does
\emph{not} multiply the $\sqrt{\sum_{t=1}^T a_t^2}$ term.  We will also use the
well-known technical Lemma~\ref{lemma:sum-of-square-roots-inverses}.

\begin{lemma}[Useful Inequality]
\label{lemma:useful}
Let $C, a_1, a_2, \dots, a_T \ge 0$. Then,
$$
\sum_{t=1}^T \min \left\{ \frac{a_t^2}{\sqrt{\sum_{i=1}^{t-1} a_i^2}}, \ C a_t \right\}
\le 3.5 C \max_{t=1,2,\dots,T} a_t + 3.5 \sqrt{\sum_{t=1}^T a_t^2} \; .
$$
\end{lemma}
\begin{proof}
Without loss of generality, we can assume that $a_t > 0$ for all $t$. Since otherwise we
can remove all $a_t = 0$ without affecting either side of the inequality. Let $M_t = \max\{a_1, a_2, \dots, a_t\}$ and $M_0 = 0$.
We prove that for any $\alpha > 1$
$$
\min\left\{ \frac{a_t^2}{\sqrt{\sum_{i=1}^{t-1} a_i^2}}, C a_t \right\}
\le 2 \sqrt{1+\alpha^2} \left( \sqrt{\sum_{i=1}^t a_i^2} - \sqrt{\sum_{i=1}^{t-1} a_i^2} \right) + \frac{C\alpha( M_t  - M_{t-1})}{\alpha - 1}
$$
from which the inequality follows by summing over $t=1,2,\dots,T$ and choosing $\alpha = \sqrt{2}$.
The inequality follows by case analysis. If $a_t^2 \le \alpha^2 \sum_{i=1}^{t-1} a_i^2$, we have
\begin{multline*}
\min\left\{ \frac{a_t^2}{\sqrt{\sum_{i=1}^{t-1} a_i^2}}, C a_t \right\}
\le \frac{a_t^2}{\sqrt{\sum_{i=1}^{t-1} a_i^2}}
= \frac{a_t^2}{\sqrt{\frac{1}{1+\alpha^2} \left( \alpha^2 \sum_{i=1}^{t-1} a_i^2 + \sum_{i=1}^{t-1} a_i^2 \right)}} \\
\le \frac{a_t^2\sqrt{1+\alpha^2}}{\sqrt{ a_t^2 + \sum_{i=1}^{t-1} a_i^2 }}
= \frac{a_t^2\sqrt{1+\alpha^2}}{\sqrt{\sum_{i=1}^t a_i^2}}
\le 2\sqrt{1+\alpha^2} \left( \sqrt{\sum_{i=1}^t a_i^2} - \sqrt{\sum_{i=1}^{t-1} a_i^2} \right)
\end{multline*}
where we have used $x^2/\sqrt{x^2+y^2} \le 2(\sqrt{x^2+y^2} - \sqrt{y^2})$ in the last step.
On the other hand, if $a_t^2 > \alpha^2 \sum_{t=1}^{t-1} a_i^2$, we have
\begin{multline*}
\min\left\{ \frac{a_t^2}{\sqrt{\sum_{i=1}^{t-1} a_i^2}}, C a_t \right\}
\le C a_t
= C \frac{\alpha a_t  - a_t}{\alpha - 1}
\le \frac{C}{\alpha - 1} \left( \alpha a_t  - \alpha \sqrt{\sum_{i=1}^{t-1} a_i^2} \right) \\
= \frac{C\alpha}{\alpha - 1} \left( a_t  - \sqrt{\sum_{i=1}^{t-1} a_i^2} \right)
\le \frac{C\alpha}{\alpha - 1} \left( a_t  - M_{t-1} \right)
= \frac{C\alpha}{\alpha - 1} \left( M_t  - M_{t-1} \right)
\end{multline*}
where we have used that $a_t = M_t$ and $\sqrt{\sum_{i=1}^{t-1} a_i^2} \ge M_{t-1}$.
\end{proof}

>>>>

The closest algorithm to a scale-free one in the OLO literature is the
\textsc{AdaGrad FTRL} algorithm~\cite{Duchi-Hazan-Singer-2011}.  It uses a
regularizer on each coordinate of the form
\begin{equation*}
R_t(w) = R(w) \left(\delta + \sqrt{\sum_{i=1}^{t-1} \|\ell_i\|_*^2} \right).
\end{equation*}
This kind of regularizer would yield a scale-free algorithm \emph{only} for
$\delta=0$.  Unfortunately, the regret bound in~\cite{Duchi-Hazan-Singer-2011}
becomes vacuous for such setting in the unbounded case. In fact, it requires
$\delta$ to be greater than $\|\ell_t\|_*$ for all time steps $t$, requiring
knowledge of the future (see Theorem~5 in~\cite{Duchi-Hazan-Singer-2011}). In
other words, despite of its name, \textsc{AdaGrad} is not fully adaptive to the
norm of the gradient vectors. Identical considerations hold for the
\textsc{FTRL-Proximal} in \cite{McMahan-Streeter-2010,McMahan-2014}: the
scale-free setting of the learning rate is valid only in the bounded case.

One simple approach would be to use a doubling trick on $\delta$ in order to
estimate on the fly the maximum norm of the losses. Note that a naive strategy
would still fail because the initial value of $\delta$ should be data-dependent
in order to have a scale-free algorithm. Moreover, we would have to upper bound
the regret in all the rounds where the norm of the current loss is bigger than
the estimate. Finally, the algorithm would depend on an additional parameter,
the ``doubling'' power. Hence, even in the case one would prove a regret bound,
such strategy would give the feeling that \textsc{FTRL} needs to be ``fixed''
in order to obtain a scale-free algorithm.

@@

\begin{theorem}[Regret of \textsc{SOLO FTRL}]
\label{theorem:regret-solo-ftrl}
Suppose $K \subseteq V$ is a non-empty closed convex set.  Let $D = \sup_{u,v
\in K} \|u - v\|$ be its diameter with respect to a norm $\|\cdot\|$.  Suppose
that the regularizer $R:K \to \R$ is a non-negative lower semi-continuous
function that is $\lambda$-strongly convex with respect to $\|\cdot\|$. The
regret of \textsc{SOLO FTRL} satisfies
\begin{align*}
\Regret_T(u)
& \le \left( R(u) + \frac{2.75}{\lambda}\right) \sqrt{\sum_{t=1}^{T} \norm{\ell_t}_*^2}
+ 3.5 \min\left\{\frac{\sqrt{T-1}}{\lambda} , D\right\} \max_{t \le T} \|\ell_t\|_* \; .
\end{align*}
\end{theorem}

@@

\begin{corollary}[Regret Bound for Bounded Decision Sets]
\label{corollary:regret-solo-ftrl-bounded-set}
Suppose $K \subseteq V$ is a non-empty bounded closed convex set.  Suppose that
$f:K \to \R$ is a non-negative lower semi-continuous function that is
$1$-strongly convex with respect to $\|\cdot\|$. \textsc{SOLO FTRL} with
regularizer
$$
R(w) = \frac{f(w)\sqrt{2.75}}{\sqrt{\sup_{v \in K} f(v)}}
\quad \text{satisfies} \quad
\Regret_T \le 13.3 \sqrt{\sup_{v \in K} f(v) \sum_{t=1}^{T} \norm{\ell_t}_*^2} \; .
$$
\end{corollary}

@@

The proof of Theorem~\ref{theorem:regret-solo-ftrl} relies on an inequality
(Lemma~\ref{lemma:useful}).  Related and weaker inequalities were proved
in~\cite{Auer-Cesa-Bianchi-Gentile-2002} and~\cite{Jaksch-Ortner-Auer-2010}.
The main property of this inequality is that on the right-hand side $C$ does
\emph{not} multiply the $\sqrt{\sum_{t=1}^T a_t^2}$ term.  We will also use the
well-known technical Lemma~\ref{lemma:sum-of-square-roots-inverses}.

\begin{lemma}[Useful Inequality]
\label{lemma:useful}
Let $C, a_1, a_2, \dots, a_T \ge 0$. Then,
$$
\sum_{t=1}^T \min \left\{ \frac{a_t^2}{\sqrt{\sum_{i=1}^{t-1} a_i^2}}, \ C a_t \right\}
\le 3.5 \, C \max_{t=1,2,\dots,T} a_t \ + \ 3.5 \sqrt{\sum_{t=1}^T a_t^2} \; .
$$
\end{lemma}
\begin{proof}
Without loss of generality, we can assume that $a_t > 0$ for all $t$. Since otherwise we
can remove all $a_t = 0$ without affecting either side of the inequality. Let $M_t = \max\{a_1, a_2, \dots, a_t\}$ and $M_0 = 0$.
We prove that for any $\alpha > 1$
$$
\min\left\{ \frac{a_t^2}{\sqrt{\sum_{i=1}^{t-1} a_i^2}}, C a_t \right\}
\le 2 \sqrt{1+\alpha^2} \left( \sqrt{\sum_{i=1}^t a_i^2} - \sqrt{\sum_{i=1}^{t-1} a_i^2} \right) + \frac{C\alpha( M_t  - M_{t-1})}{\alpha - 1}
$$
from which the inequality follows by summing over $t=1,2,\dots,T$ and choosing $\alpha = \sqrt{2}$.
The inequality follows by case analysis. If $a_t^2 \le \alpha^2 \sum_{i=1}^{t-1} a_i^2$, we have
\begin{multline*}
\min\left\{ \frac{a_t^2}{\sqrt{\sum_{i=1}^{t-1} a_i^2}}, C a_t \right\}
\le \frac{a_t^2}{\sqrt{\sum_{i=1}^{t-1} a_i^2}}
= \frac{a_t^2}{\sqrt{\frac{1}{1+\alpha^2} \left( \alpha^2 \sum_{i=1}^{t-1} a_i^2 + \sum_{i=1}^{t-1} a_i^2 \right)}} \\
\le \frac{a_t^2\sqrt{1+\alpha^2}}{\sqrt{ a_t^2 + \sum_{i=1}^{t-1} a_i^2 }}
= \frac{a_t^2\sqrt{1+\alpha^2}}{\sqrt{\sum_{i=1}^t a_i^2}}
\le 2\sqrt{1+\alpha^2} \left( \sqrt{\sum_{i=1}^t a_i^2} - \sqrt{\sum_{i=1}^{t-1} a_i^2} \right)
\end{multline*}
where we have used $x^2/\sqrt{x^2+y^2} \le 2(\sqrt{x^2+y^2} - \sqrt{y^2})$ in the last step.
On the other hand, if $a_t^2 > \alpha^2 \sum_{t=1}^{t-1} a_i^2$, we have
\begin{multline*}
\min\left\{ \frac{a_t^2}{\sqrt{\sum_{i=1}^{t-1} a_i^2}}, \ C a_t \right\}
\le C a_t
= C \frac{\alpha a_t  - a_t}{\alpha - 1}
\le \frac{C}{\alpha - 1} \left( \alpha a_t  - \alpha \sqrt{\sum_{i=1}^{t-1} a_i^2} \right) \\
= \frac{C\alpha}{\alpha - 1} \left( a_t  - \sqrt{\sum_{i=1}^{t-1} a_i^2} \right)
\le \frac{C\alpha}{\alpha - 1} \left( a_t  - M_{t-1} \right)
= \frac{C\alpha}{\alpha - 1} \left( M_t  - M_{t-1} \right)
\end{multline*}
where we have used that $a_t = M_t$ and $\sqrt{\sum_{i=1}^{t-1} a_i^2} \ge M_{t-1}$.
\end{proof}
